Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-06 appendix.tex
major revision underway, update to include global correction in numerical sims

The `global correction' terms do not change the qualitative dependence
of the fundamental VSI growth rates on the thermal relaxation
timescale. The difference is negligble for $\beta\lesssim 1$ (i.e.
fast thermal relaxation). The critical relaxation timescale for the
fundamental VSI is unaffected by these global radients.

>>>>

The global correction terms do not change the qualitative dependence
of the fundamental VSI growth rates on the thermal relaxation
timescale. The difference is negligble for $\beta\lesssim 1$ (i.e.
fast thermal relaxation). The critical relaxation timescale for the
fundamental VSI is unaffected by these global radial radients.

### 2015-02-06 application.tex
abstract added, intro left to do
major revision underway, update to include global correction in numerical sims

\section{Application to protoplanetary disks}\label{application} 
As an application of our results, we estimate where in a 
protoplanetary disk do thermodynamic conditions allow the VSI to 
operate. Specifically, we consider the fundamental isothermal VSI, 
for which we can use the criterion Eq. \ref{iso_vsi_cond}.

In our disk models we have adopted a simple thermal relaxation 
function in the energy equation. To connect our model with a realistic
disk, we consider a more physically-motivated form of the energy
equation with radiative cooling 
\begin{align}\label{real_energy}
\rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}
where the temperature $T$ is defined through $P=\mathcal{R}\rho
T/\mu$, where $\mathcal{R}$ is the ideal gas constant and $\mu$ is
the mean molecular weight. In Eq. \ref{real_energy} $S$ is the
specific entropy,
\begin{align}
  S \equiv c_v\ln{\left(\frac{p}{\rho^{\gamma}}\right)} 
\end{align}
where $c_v \equiv \left(\gamma-1\right)^{-1}\mathcal{R}/\mu$ is the
specific heat capcity at constant volume, $D/Dt$ is the convective
derivative, $\bm{F}$ is the radiative flux, and $Q_+$ is a heating
term such that an equilibrium solution exists when $D/Dt=0$.

Our aim is to relate the right hand side (RHS) of Eq. \ref{real_energy} to
the thermal relaxation time $t_c$ as used in our linear
calculations. To do so, we first consider separately perturbations
with radial lengthscales $L\equiv 2\pi/k_x\ll l_\mathrm{rad}$ and 
$L\gg l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity. 
%need to comment on kappa_d(rho,T)

\subsection{Newtonian cooling}\label{newton_cool}
For $L\ll l_\mathrm{rad}$ the radiative cooling is given by {\bf (reference?)}
\begin{align}
\nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
\end{align}   
where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
heating term 
\begin{align}
  Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
\end{align}
Then Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
  \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right). 
\end{align}
When linearized, the last term on the RHS becomes
\begin{align}
  -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
    \frac{P}{\rho}\delta\rho\right). 
\end{align}
By comparing this expression with the linearized version of the energy
equation used in our calculations (Eq. \ref{energy_eq}), we identify 
\begin{align}\label{tc_newton_cool} 
  t_\mathrm{cool} = \frac{c_v}{16\sigma_s\kappa_dT^3}
\end{align}
as the thermal relaxation timescale for perturbation lengthscales
$L\ll l_\mathrm{rad}$.

\subsection{Radiative diffusion}
For perturbation lengthscales $L\gg l_\mathrm{rad}$, let us assume that
temperature fluctuations are smoothed out by radiative diffusion. To
model this in linear theory, we choose the heating term such that
\begin{align}
  Q_+ - \nabla\cdot\bm{F} \equiv \nabla\cdot\left[k_\mathrm{rad}\nabla
    \left(T-T_{t=0}\right)\right], 
\end{align} 
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly. Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} + \frac{P}{\rho T
    c_v}\nabla\cdot\left[k_\mathrm{rad}\nabla\left(T-T_{t=0}\right)\right].     
\end{align}
When linearized, the last term on the RHS becomes
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right). 
\end{align}
This, of course, fundamentally changes the linear problem by
introducing higher order derivatives. We defer a proper analysis of
this problem to a follow-up study. Here, we are interested in
order-of-magnitude estimates for perturbations with radial
lengthscales much smaller than vertical lengthscales. We thus proceed
by assuming vertical derivatives can be neglected in
Eq. \ref{diff_cool_proper} and write  
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    c_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
% we are interested in radially-thin perturbations  
where $\eta=k_\mathrm{rad}/\rho c_v$ is the diffusion coefficient and
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}. 
\end{align}
From Eq. \ref{diff_cool_approx} we identify the thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}\equiv \frac{\Omega_k^{-1}}{\hat{\eta}\khat^2}, 
\end{align}
where $\hat{\eta} = \eta/H_\mathrm{iso}^2\Omega_k$ is the 
dimensionless diffusion coefficient.

@@

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $\zeta\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
\beta =
\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)\zeta^{-57/14}\left(\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}\zeta^{33/7}+\frac{1}{\khat^2}\right).    
\end{align}
We also evaluate the criterion Eq. \ref{iso_vsi_cond} for the fiducial
disk model,  
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}\zeta^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is a neccessary condition for the
fundamental isothermal VSI when $\khat\gg 1$. For $\khat\to0$, this
becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
\S\ref{iso_vsi_beta_crit}).

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn} and
Eq. \ref{bcrit_mmsn} for the Minimum Mass Solar Nebula (MMSN) by setting
the scalings $\hat{T}=\hat{\Sigma}=\hat{\kappa}_d=1$, with $\mu =
2.33$ and $\gamma=1.4$. The figure suggests that the fundamental
isothermal VSI operates in the outer parts of the MMSN at radii on
the order of tens of AU. However, the thermal relaxation time $\beta$
is not much shorter than $\beta_\mathrm{crit}$, so the VSI may be 
weak.

@@

\subsection{Numerical example}\label{mmsn_example}
Consider  the MMSN at $r=50\mathrm{AU}$. Eq. \ref{mmsn_epsilon} gives
$\epsilon \simeq 
0.067$ which implies, using Eq. \ref{bcrit_mmsn}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.072. 
\end{align*} 
To estimate a thermal relaxation timescale, we set $\khat=10$ 
in Eq. \ref{beta_mmsn} to obtain 
\begin{align*}
  \beta \simeq 0.049.
\end{align*}
(In fact, Fig. \ref{mmsn_bcrit_bcool} shows that $\beta$  is not
sensitive to $\khat$ at $r=50\mathrm{AU}$ for $\khat\gg1$.)
The disk is unstable to the fundamental isothermal VSI, but not
strongly.

@@

\subsubsection{Vertically-depedent thermal relaxation} 
Note that for a vertically isothermal disk and opacity $\kappa_d(T)$, 
we can restore the $z$ dependence in $\beta$ (due to the $\rho$
dependence in  $\eta$, see Eq. \ref{eta_def}) by making the transformation
$\beta(\khat)\to\beta(\khat/\sqrt{2\pi}\hat{\rho})\equiv 
\tilde{\beta}(\khat,z)$ in  Eq. \ref{real_beta}, where
$\hat{\rho}=\rho/\rho(z=0)$.  We have have repeated the example
calculation above with $\beta\to\tilde{\beta}$. We find growth rates
were only slightly reduced at low $\khat$ and eigenfunctions are
similar.

>>>>

\section{Application to protoplanetary disks}\label{application} 
As an application of our results, we estimate where in a 
protoplanetary disk do thermodynamic conditions allow the VSI to 
operate. Specifically, we consider the fundamental VSI, 
for which we can use the criterion Eq. \ref{iso_vsi_cond}.

In our disk models we have adopted a simple thermal relaxation 
function in the energy equation. To connect our model with a realistic
disk, we consider a more physically-motivated form of the energy
equation with radiative cooling 
\begin{align}\label{real_energy}
\rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}
where the temperature $T$ is defined through $P=\mathcal{R}\rho
T/\mu$, $\mathcal{R}$ is the ideal gas constant and $\mu$ is
the mean molecular weight, % In Eq. \ref{real_energy} $S$ is the
% specific entropy,
% \begin{align}
%   S \equiv c_v\ln{\left(\frac{p}{\rho^{\gamma}}\right)} 
% \end{align}
% where $c_v \equiv \left(\gamma-1\right)^{-1}\mathcal{R}/\mu$ is the
% specific heat capcity at constant volume
$D/Dt$ is the convective
derivative, $\bm{F}$ is the radiative flux, and $Q_+$ is a heating
term such that an equilibrium solution exists when $D/Dt=0$.

Our aim is to relate the right hand side (RHS) of Eq. \ref{real_energy} to
the thermal relaxation time $t_c$ as used in our linear
calculations. To do so, we first consider separately perturbations
with radial lengthscales $l\equiv 2\pi/k_x\ll l_\mathrm{rad}$ and 
$l\gg l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity. 
%need to comment on kappa_d(rho,T)

\subsection{Newtonian cooling}\label{newton_cool}
For $l\ll l_\mathrm{rad}$ the radiative cooling is given by {\bf (reference?)}
\begin{align}
\nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
\end{align}   
where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
heating term 
\begin{align}
  Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
\end{align}
Then Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
  \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right). 
\end{align}
When linearized, the last term on the RHS becomes
\begin{align}
  -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
    \frac{P}{\rho}\delta\rho\right). 
\end{align}
By comparing this expression with the linearized version of the energy
equation used in our calculations (Eq. \ref{energy_eq}), we identify 
\begin{align}\label{tc_newton_cool} 
  t_\mathrm{cool} = \frac{c_v}{16\sigma_s\kappa_dT^3}
\end{align}
as the thermal relaxation timescale for perturbation lengthscales
$L\ll l_\mathrm{rad}$.

\subsection{Radiative diffusion}
For perturbation lengthscales $l\gg l_\mathrm{rad}$, we assume 
temperature fluctuations are smoothed out by radiative diffusion. To
model this in linear theory, we choose the heating term such that
\begin{align}
  Q_+ - \nabla\cdot\bm{F} \equiv \nabla\cdot\left[k_\mathrm{rad}\nabla
    \left(T-T_{t=0}\right)\right], 
\end{align} 
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly. Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} + \frac{P}{\rho T
    c_v}\nabla\cdot\left[k_\mathrm{rad}\nabla\left(T-T_{t=0}\right)\right].     
\end{align}
When linearized, the last term on the RHS becomes
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right). 
\end{align}
This, of course, fundamentally changes the linear problem by
introducing higher order derivatives. We defer a proper analysis of
this problem to a follow-up study. Here, we are interested in
order-of-magnitude estimates for perturbations with radial
lengthscales much smaller than vertical lengthscales. We thus proceed
by assuming vertical derivatives can be neglected in
Eq. \ref{diff_cool_proper} and write  
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    c_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
% we are interested in radially-thin perturbations  
where $\eta=k_\mathrm{rad}/\rho c_v$ is the diffusion coefficient and
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}. 
\end{align}
From Eq. \ref{diff_cool_approx} we identify the thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}\equiv \frac{\Omega_k^{-1}}{\hat{\eta}\khat^2}, 
\end{align}
where $\hat{\eta} = \eta/H_\mathrm{iso}^2\Omega_k$ is the 
dimensionless diffusion coefficient.

@@

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $\zeta\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
\beta =
\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)\zeta^{-57/14}\left(\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}\zeta^{33/7}+\frac{1}{\khat^2}\right).    
\end{align}
We also evaluate the criterion Eq. \ref{iso_vsi_cond} for the fiducial
disk model,  
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}\zeta^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is neccessary for the
fundamental VSI when $\khat\gg 1$. For $\khat\to0$, this
becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
\S\ref{iso_vsi_beta_crit}).

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn} and
Eq. \ref{bcrit_mmsn} for the Minimum Mass Solar Nebula (MMSN) by setting
the scalings $\hat{T}=\hat{\Sigma}=\hat{\kappa}_d=1$, with $\mu =
2.33$ and $\gamma=1.4$. The figure suggests that the fundamental VSI
operates in the outer parts of the MMSN at radii on 
the order of tens of AU. However, the thermal relaxation time $\beta$
is not much shorter than $\beta_\mathrm{crit}$, so the VSI may be 
weak.

@@

\subsection{Numerical example}\label{mmsn_example}
Consider  the MMSN at $r=50\mathrm{AU}$. Eq. \ref{mmsn_epsilon} gives
$\epsilon \simeq 
0.067$ which implies, using Eq. \ref{bcrit_mmsn}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.072. 
\end{align*} 
To estimate a thermal relaxation timescale, we set $\khat=10$ 
in Eq. \ref{beta_mmsn} to obtain 
\begin{align*}
  \beta \simeq 0.049.
\end{align*}
(In fact, Fig. \ref{mmsn_bcrit_bcool} shows that $\beta$  is not
sensitive to $\khat$ at $r=50\mathrm{AU}$ for $\khat\gg1$.)
The MMSN is unstable to the fundamental VSI, but not
strongly.

@@

\subsubsection{Vertically-depedent thermal relaxation} 
Note that for a vertically isothermal disk and opacity $\kappa_d(T)$, 
we can restore the $z$ dependence in $\beta$ (due to the $\rho$
dependence in  $\eta$, see Eq. \ref{eta_def}) by making the transformation
$\beta(\khat)\to\beta(\khat/\sqrt{2\pi}\hat{\rho})\equiv 
\tilde{\beta}(\khat,z)$ in  Eq. \ref{real_beta}, where
$\hat{\rho}=\rho/\rho(z=0)$.  We have have repeated the example
calculation above with $\beta\to\tilde{\beta}$. We find growth rates
were reduced, but at most by $\sim 15\%$ at $\khat=1$. Eigenfunctions
were similar.

%at most by 15%

### 2015-02-06 numerical.tex
major revision underway, update to include global correction in numerical sims

Fig. \ref{relax_growth_num} shows the growth rates of the fundamental  
mode as a function of $\khat$. This plot is largely consistent with  
Fig. \ref{relax_disp_fig} except for $\khat\lesssim 1$
where the low-frequency approximation fails. (The local model not valid
for $\khat\ll1$ anyway). For $\khat\gtrsim 10$ introducing thermal
relaxation rapidly stabilizes the fundamental mode.

@@

\subsubsection{Dependence on disk parameters}
We test our estimate for the critical relaxation time
$\beta_\mathrm{crit}$ (Eq. \ref{iso_vsi_cond}). We use the above disk
parameters as a fiducial case and vary $q\in[-0.2,-1.2]$,
$\gamma\in[1.2,2.0]$ and $\epsilon\in[]$
separately. The numerically-obtained $\beta_\mathrm{crit}$ is shown in
Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/bcrit_compare_q.ps} 
  \caption{Dependence of the upper limit to the thermal relaxation timescale
    $\beta_\mathrm{crit}$ for the fundamental VSI mode on disk
    parameters. The fiducial parameters are $\Gamma=1.011$,
    $\gamma=1.4$ and $(p,q,\epsilon)=(-1.5,-1,0.05)$. Top: varying
    $q\in[-1.2,-0.2]$; middle: varying $\gamma\in[1.2,2.0]$; bottom:
    varying $\epsilon\in[]$. The perturbation wavenumber is
    $\khat=20$. {\bf waiting for high resolution}
    \label{bcrit_compare}}  
\end{figure}

@@

\subsubsection{$\gamma=\Gamma \neq 1$, $N_z^2\equiv 0$}
Consider first the case $\gamma=1.4$ so that there is no vertical
entropy gradient. We plot in Fig. the growth rates for the
fundamental VSI for $\beta\in[10^{-2},10^{2}]$. We find growth rates
do not reach zero, implying the VSI operates regardless of the thermal
relaxation timescale if $\gamma=\Gamma$. This result is consistent
with \cite{nelson13}.

We also note that growth rates for $\beta = 10^{-2}$ and
$\beta=10^{2}$ behave similarly. This is not surprising because
the linearized energy equation (Eq. \ref{lin_energy}) implies that 
\begin{align}
  W = \frac{\left(\Gamma^{-1} -
      \ii\hat{\sigma}\beta\right)}{\left(1-\ii\hat{\sigma}\beta\right)}
  Q \text{ for } \gamma\equiv \Gamma.   
\end{align}
Thus, $W \to Q/\Gamma$ as $\beta\to0$ and $W\to Q$ as
$\beta\to\infty$. Since $\Gamma=O(1)$, these two limits are expected
to produce qualitatively similar results. Interestingly,
Fig. \ref{growth_vnoniso1} show that growth rates 
maximize for $\beta=O(1)$. However, all growth rates are similar to
order-of-magnitude for $0.1\lesssim\khat\lesssim 100$.

@@

\subsubsection{$\gamma\neq\Gamma \neq 1$, $N_z^2\geq0$}
%gamma = 2.0, Gamma = 1.4, vary bcool 
Finally, we introduce buoyancy in a vertically non-isothermal disk by
setting $\gamma=2$. This scenario was not considered by  
\cite{nelson13}, but is similar in spirit with vertically isothermal 
disks evolved with $\gamma>1$, as considered in the previous section. 
Fig. \ref{growth_vnoniso2} plots the fundamental VSI growth rates for
this case. We observe the same trend as for vertically isothermal
disks with $N_z^2\geq 0$ (Fig. \ref{relax_growth_num}), namely growth
rates reach zero at finite $\khat$.

>>>>

Fig. \ref{relax_growth_num} shows the growth rates of the fundamental  
mode as a function of $\khat$. This plot is largely consistent with  
Fig. \ref{relax_disp_fig} except for $\khat\lesssim 1$
where the low-frequency approximation fails. (In any case, the local model not valid
for $\khat\ll1$.) For $\khat\gtrsim 10$ introducing thermal
relaxation rapidly stabilizes the fundamental mode.

@@

\subsection{Critical thermal relaxation timescale}
We test our estimate for the maximum thermal relaxation timescale 
$\beta_\mathrm{crit}$ for the fundamental
VSI (Eq. \ref{iso_vsi_cond}). We use the previous disk
parameters as a fiducial case and vary $q\in[-0.2,-1.2]$,
$\gamma\in[1.2,2.0]$ and $\epsilon\in[]$
separately. The numerically-obtained $\beta_\mathrm{crit}$ is shown in
Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.

We find Fig. \ref{bcrit_compare} generally agrees with
Eq. \ref{iso_vsi_cond}.  The match is better for smaller
$\beta_\mathrm{crit}$. There is a noticeable difference as
$\gamma\to1$ because the  derivation of $\beta_\mathrm{crit}$ assumed
$\gamma>1$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0cm
  0cm]{figures/bcrit_compare_q.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.8cm]{figures/bcrit_compare_g.ps} 
  \caption{Dependence of the upper limit to the thermal relaxation timescale
    $\beta_\mathrm{crit}$ for the fundamental VSI mode on disk
    parameters. The fiducial parameters are $\Gamma=1.011$,
    $\gamma=1.4$ and $(p,q,\epsilon)=(-1.5,-1,0.05)$. Top: varying
    $q\in[-1.2,-0.2]$; middle: varying $\gamma\in[1.2,2.0]$; bottom:
    varying $\epsilon\in[]$. The perturbation wavenumber is
    $\khat=20$. {\bf still need variable epsilon case}
    \label{bcrit_compare}}  
\end{figure}

@@

\subsubsection{$\gamma=\Gamma \neq 1$, $N_z^2\equiv 0$}
Consider first the case $\gamma=1.4$ so that there is no vertical
entropy gradient. We plot in Fig. the growth rates for the
fundamental VSI for $\beta\in[10^{-2},10^{2}]$. We find growth rates
do not reach zero, implying instability regardless of the thermal
relaxation timescale if $\gamma=\Gamma$. This result is consistent
with \cite{nelson13}.  Note that for $\beta\to\infty$ (adiabatic
perturbations) and $\gamma=\Gamma$, the disk can be unstable according
to the Solberg-Hoiland criterion  Eq. \ref{solberg2}.

Growth rates for $\beta = 10^{-2}$ and
$\beta=10^{2}$ behave similarly. This is not surprising because
the linearized energy equation (Eq. \ref{lin_energy}) implies that 
\begin{align}
  W = \frac{\left(\Gamma^{-1} -
      \ii\hat{\sigma}\beta\right)}{\left(1-\ii\hat{\sigma}\beta\right)}
  Q \text{ for } \gamma\equiv \Gamma.   
\end{align}
Thus, $W \to Q/\Gamma$ as $\beta\to0$ and $W\to Q$ as
$\beta\to\infty$. Since $\Gamma=O(1)$, these two limits are expected
to produce qualitatively similar results. Interestingly,
Fig. \ref{growth_vnoniso1} show that growth rates 
maximize for $\beta=O(1)$. However, all growth rates are similar to
order-of-magnitude for $0.1\lesssim\khat\lesssim 100$.

@@

\subsubsection{$\gamma\neq\Gamma \neq 1$, $N_z^2\geq0$}
%gamma = 2.0, Gamma = 1.4, vary bcool 
Finally, we introduce buoyancy in a vertically non-isothermal disk by
setting $\gamma=2$. This scenario was not considered by  
\cite{nelson13}, but is similar in spirit with vertically isothermal 
disks evolved with $\gamma>1$, as considered in the previous section. 
Fig. \ref{growth_vnoniso2} plots the fundamental VSI growth rates for
this case. We observe the same trend as for vertically isothermal
disks with $N_z^2\geq 0$ (Fig. \ref{relax_growth_num}), namely growth
rates reach zero at finite $\khat$. From Fig.\ref{growth_vnoniso2} we
expect the VSI to be strongly stabilized even for small, finite
thermal relaxation times.

### 2015-02-06 paper.tex
major revision underway, update to include global correction in numerical sims

However, MHD turbulence may not be applicable for all 
accretion disks. For example, there may be regions in protoplanetary
disks --- called dead zones ---  which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent state-of-the-art

%in this respect purely hydrodynamical turbulence should be

%stuff about vsi

>>>>

However, MHD turbulence may not be applicable for all  
accretion disks. For example, there may be regions in protoplanetary  
disks which do not support the MRI \citep{gammie96,landry13}. In fact,
state-of-the-art  numerical simulations suggest portions of PPDs
remain laminar. %transport by wind, large-scale field 
% cite bai, lesur,
% gressel

In this respect, purely hydrodynamical turbulence remains of interest
in the context of PPDs. One possibiliy is the vertical shear
instability (VSI), which has recently been demonstrated through
hydrodynamical simulations \citep{nelson13} %and kley simulations

%stuff about vsi 
The VSI is an axisymmetric instability in three-dimensional (3D) disks
for which the rotation frequency $\Omega$ depends on the height from
the disk midplane.
